*DCVtestkit*: a R package for linearity assessment and visualisation of multiple curves

### 2022-11-09 Manuscript/manuscript.qmd
Update the introduction to be more general.

Linearity assessment is one of the criteria used to evaluate the precision of a quantitative measurement procedure as seen in calibration studies from @rodríguez1993 and drug analysis from @needleman1990. In metabolomics/lipidomics analysis, dilution curves are plotted for each molecule measured from quality control sample, generated by pooling equal aliquots from each study sample. Molecules that exhibits a non-linear relationship are rejected before further statistical analysis, as their quantification in different sample groups might be compromised. Dilution curves are usually plotted using general-purpose software like Excel with their corresponding Pearson Correlation Coefficient value. A threshold value of 0.8 is used to balance the risk of accepting or removing non-linear signals related to relevant molecules.

However, @sonnergaard2006 warned that the Pearson Correlation Coefficient is not an effective standalone numeric parameter to estimate linearity. While researchers have created other metrics for linearity evaluation, these metrics are rarely implemented in most general-purpose software. Furthermore, today's metabolomics/lipidomics workflow can measure hundreds of molecules. Having the analyst to individually plot numerous curves to check for linearity is time consuming.

R package, *DCVtestkit* addresses these issues by assisting analysts, to plot dilution curves for many molecules easily with additional metrics, other than the Pearson Correlation Coefficient, that better describe the curve's shape. It also provides an interactive viewer for analysts to group, filter and sort the plots, allowing them to look at the relevant ones, such as the saturated dilution curves, and identify problematic molecules quickly.

@@

Supplementary Figure 1 depicts the workflow of *DCVtestkit*. The workflow starts with two tables: Transition Signal Data, containing transition signals (y-axis for dilution curve) for each sample and Dilution Annotation, containing dilution curve related information, such as concentration (x-axis for dilution curve) and dilution batches. Using a common column Sample Name, the two tables can be merged into one table (Dilution Table) via `create_dilution_table`.

@@

To verify if a quantitative analytical test method is reliable, it is important to check for linearity. A linear response in quality control dilution curves is required in metabolomics/lipidomics to assess the molecules' suitability for further analysis. However, there are few software that can analyse dilution curves efficiently. R package, *DCVtestkit*, rectifies this by plotting of many dilution curves quickly by automation and reporting alternative statistics, other than the Pearson Correlation Coefficient, to better describe the shape of dilution curves. It also provides an interactive trellis plot for exploratory data analysis. It is available on GitHub <https://github.com/SLINGhub/DCVtestkit> while the documentation and tutorials can be accessed from <https://slinghub.github.io/DCVtestkit>.

>>>>

Linearity assessment, as summarised by @paulson1995, is a standard protocol to verify if an instrument or experimental method is in working condition, especially if the method is used to report quantitative results. It is applied in many fields in science such as calibration/dilution studies (@rodríguez1993 and @sands2021) and assay development (@ross2003 and @hsieh2008). During analysis, curves are plotted individually with a Pearson Correlation Coefficient value for each measurement using general-purpose software like Excel.

However, as instrument or experimental method are becoming more high throughput, many measurements can be done at a short time. Having the analyst to individually plot numerous curves to check for linearity is time-consuming. Furthermore, @sonnergaard2006 warns that the Pearson Correlation Coefficient is not an effective standalone numeric parameter to estimate linearity. While researchers have created other metrics for linearity evaluation, these metrics are rarely implemented in most general-purpose software.

R package, *DCVtestkit* addresses these issues by assisting analysts, to plot curves from many experiments easily with additional metrics, other than the Pearson Correlation Coefficient, that better describe the curve's shape. It also provides an interactive viewer for analysts to group, filter and sort the plots, allowing them to look at problematic ones, such as saturated curves.

@@

Using the analysis of dilution curves in meatbolomic/lipidomic study as a running example, Supplementary Figure 1 depicts the workflow of *DCVtestkit*. The workflow starts with two tables: Transition Signal Data, containing transition signals (y-axis for dilution curve) for each sample and Dilution Annotation, containing dilution curve related information, such as concentration (x-axis for dilution curve) and dilution batches. Using a common column Sample Name, the two tables can be merged into one table (Dilution Table) via `create_dilution_table`.

@@

To verify if a quantitative analytical test method is reliable, it is important to check for linearity. A linear response in quality control dilution curves is required in metabolomics/lipidomics to assess the molecules' suitability for further analysis. However, there are few software that can analyse dilution curves efficiently.

R package, *DCVtestkit*, rectifies this by plotting of many curves quickly by automation and reporting alternative statistics, other than the Pearson Correlation Coefficient, to better describe the shape of curves. It also provides an interactive trellis plot for exploratory data analysis. It is available on GitHub <https://github.com/SLINGhub/DCVtestkit> while the documentation and tutorials can be accessed from <https://slinghub.github.io/DCVtestkit>.
